---
title: "Weather Events Analysis in the United States"
output: 
    html_document:
        keep_md: true
---

```{r setup, include=FALSE}
library(dplyr);library(reshape2);library(ggplot2)
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis
The major storms and weather events characteristics is explored to find the particular events that cause the most public health issues and economic consequences. In this report, the public health issues are attributed to fatalities and injuries that occurred during these events. The economic consequence combines both the property damage and crop damage. Due to untidy Event Type data, cleaning and generalizations are made on similar categories. The similar Event types (EVTYPE) have their fatalities summed up. The same is done with injuries, property damage cost and crop damage cost. The results are ploted using a stacked bar plot.

## Introduction
 [[1]](#refer)Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

The project will focus on two questions;
    1- Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
    2- Across the United States, which types of events have the greatest economic consequences?

## Data Processing
###    1- Data Loading
The data is retrive from the [link here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2), using R to download and read.
See [Apendix](#apendix) for [code repository](#code).
```{r cache = TRUE}
source('init.R') #Initialising file to load librarys and functions
#List of required libraries in the apendix
loadData() ## load the Data
#Function def in apendix
```
###     2- Basic exploratory analysis
```{r dataSet, cache=TRUE}
names(dataSet)
# tranform to lower case for standardization
dataSet$EVTYPE <- tolower(dataSet$EVTYPE)
# Subset data to include relevant columns   "STATE__"   "COUNTY"    "COUNTYNAME"    "STATE"     "EVTYPE"    "FATALITIES"
#                                           "INJURIES"  "PROPDMG"   "PROPDMGEXP"    "CROPDMG"   "CROPDMGEXP"
dataSub <- dataSet[,names(dataSet)[c(1,5,6,7,8,23,24,25,26,27,28)]]
head(dataSub,5)
tail(dataSub,5)
str(dataSub)
```

```{r, eval=FALSE}
levels(dataSub$EVTYPE)
```
Different EVTYPE [levels](#evtype) in [apendix](#apendix)


### 4- Data Cleaning
The event types EVTYPE multiple whitespaces inbetween that would be clean. Later spelling mistake and same category events would be replace with a group name. This would be done on only most appearing event types to minimize time spent on cleaning
```{r dataSub, cache=TRUE}

dataSub$EVTYPE <-gsub("(^[[:space:]]+|[[:space:]]+$)", "", dataSub$EVTYPE)
dataSub$EVTYPE <-gsub("\\s+", " ", dataSub$EVTYPE)

# rename all winds & wrongly spelt wind to wind
dataSub$EVTYPE <- gsub("wnd","wind",dataSub$EVTYPE, fixed = TRUE)
dataSub$EVTYPE <- gsub("winds","wind",dataSub$EVTYPE, fixed = TRUE)
dataSub$EVTYPE <- gsub("wins","wind",dataSub$EVTYPE, fixed = TRUE)

# rename all tstm & wrongly spelt thunderstorm to thunderstorm
dataSub$EVTYPE <- gsub("tstm","thunderstorm",dataSub$EVTYPE, fixed = TRUE)
dataSub$EVTYPE <- gsub("thun[a-z]+m","thunderstorm",dataSub$EVTYPE, fixed = TRUE)
dataSub$EVTYPE <- gsub("thunderstormwinds","thunderstorm w",dataSub$EVTYPE, fixed = TRUE)
dataSub$EVTYPE <- gsub("thunderstormw","thunderstorm w",dataSub$EVTYPE, fixed = TRUE)
 # replace events with a group name
dataSub[grepl("thunderstorm wind",dataSub$EVTYPE, fixed = TRUE),]$EVTYPE <- "thunderstorm w"
dataSub[grepl("thun.+ wind",dataSub$EVTYPE),]$EVTYPE <- "thunderstorm w"
dataSub[grepl("thunderstorm w inds",dataSub$EVTYPE),]$EVTYPE <- "thunderstorm w"
dataSub[grepl("tornado",dataSub$EVTYPE),]$EVTYPE <- "tornado"
dataSub[grepl("torndao",dataSub$EVTYPE),]$EVTYPE <- "tornado"
dataSub[grepl("avalance",dataSub$EVTYPE),]$EVTYPE <- "avalanche"
dataSub[grepl("wind", dataSub$EVTYPE),]$EVTYPE <- "other wind"
dataSub[grepl(".*flo*d.*",dataSub$EVTYPE),]$EVTYPE <- "flood"
dataSub[grepl(".*cold.*|.*snow.*|.*ice.*|.*winter.*|.*freez.*|.*hail.*|.*blizz.*|.*frost.*",dataSub$EVTYPE),]$EVTYPE <- "snow related"
heatRel <- dataSub[grepl(".*heat.*|.*hypo.*",dataSub$EVTYPE),]$EVTYPE
dataSub[grepl(".*heat.*|.*hypo.*",dataSub$EVTYPE),]$EVTYPE <- "heat related"
dataSub[grepl(".*rain.*",dataSub$EVTYPE),]$EVTYPE <- "rains"
dataSub[grepl(".*lightning.*",dataSub$EVTYPE),]$EVTYPE <- "lightning"
dataSub[grepl("rip current",dataSub$EVTYPE),]$EVTYPE <- "rip currents"
dataSub[grepl(".*fire.*",dataSub$EVTYPE),]$EVTYPE <- "fire related"
dataSub[grepl(".*hurricane.*",dataSub$EVTYPE),]$EVTYPE <- "hurricane"
dataSub[grepl(".*surf.*",dataSub$EVTYPE),]$EVTYPE <- "surf related"
dataSub[grepl("thunderstorm w",dataSub$EVTYPE),]$EVTYPE <- "thunderstorm wind"

dataSub$EVTYPE <- as.factor(dataSub$EVTYPE)
# all other winds

```

## Results
The public health issues include fatalities and injuries. The total fatalities and injuries for category of EVTYPE is calculated.
``` {r}
fatalitiesByType <- aggregate(FATALITIES ~ EVTYPE, dataSub, sum)
fatalitiesNonZero <- fatalitiesByType %>% filter(FATALITIES >0) %>% arrange(desc(FATALITIES))
injuriesByType <- aggregate(INJURIES ~ EVTYPE, dataSub, sum)
injuriesNonZero <- injuriesByType %>% filter(INJURIES >0) %>% arrange(desc(INJURIES))
#names(dataSet)
healthIssuesByType <- merge(fatalitiesByType,injuriesByType,by='EVTYPE')
healthIssuesByType <- healthIssuesByType %>% mutate(HEALTHISSUES = FATALITIES + INJURIES) %>% filter(HEALTHISSUES >0) %>% arrange(desc(HEALTHISSUES))
```
The top 20 EVTYPE is taken as we need the EVTYPE with the most health issues
```{r}
temp <- head(healthIssuesByType,20)
meltedIssuesByType <- melt(temp, measure.vars = c("FATALITIES","INJURIES"), id = "EVTYPE")
```
final results ploted on a bar chart.
It is seen that Tornados have the highest public health issues including fatalities. THis is followed by [heat related weather conditions](#heatrelated) then Floods and Thunderstorm winds with close marks.
```{r}
p<-ggplot(data=arrange(meltedIssuesByType,desc(value)), aes(x=reorder(EVTYPE, -value), y=value,fill=variable)) +
    geom_bar(stat="identity")+ggtitle("Public Health Issues by Weather Event Type")+ylab("Number of Issues")+xlab("Weather Events")

p + coord_flip()
```

The economic consequences will be measured by property damage and crop damage. The total cost of property damage and crop damage for category of EVTYPE is calculated.
``` {r cache = TRUE}
propertyDMGByType <- aggregate(PROPDMG ~ EVTYPE, dataSub, sum)
propertyDMGNonZero <- propertyDMGByType %>% filter(PROPDMG >0) %>% arrange(desc(PROPDMG))
cropDMGByType <- aggregate(CROPDMG ~ EVTYPE, dataSub, sum)
cropDMGNonZero <- cropDMGByType %>% filter(CROPDMG >0) %>% arrange(desc(CROPDMG))
#names(dataSet)
econDMGByType <- merge(propertyDMGByType,cropDMGByType,by='EVTYPE')
econDMGByType <- econDMGByType %>% mutate(ECONDMG = PROPDMG + CROPDMG) %>% filter(ECONDMG >0) %>% arrange(desc(ECONDMG))
```
The top 20 EVTYPE is taken as we need the EVTYPE with the economic consequences.
```{r}
temp <- head(econDMGByType,20)
meltedDMGByType <- melt(temp, measure.vars = c("PROPDMG","CROPDMG"), id = "EVTYPE")
```
final results ploted on a bar chart.
Again Tornados have the highest economic consequence. This is followed by Floods and Thunderstorm winds. Heat related weather almost goes to 0% of economic consequence. Considering the crop damage (bars in blue), snow related weather conditions cause the most economic consequence.
```{r}
p<-ggplot(data=arrange(meltedDMGByType,desc(value)), aes(x=reorder(EVTYPE, -value), y=value,fill=variable)) +
    geom_bar(stat="identity")+ggtitle("Economic Consequence by Weather Event Type")+ylab("Cost")+xlab("Weather Events")

p + coord_flip()
```

## Conclusion
The database was explored and results suggest tornado cause both the highest public health issues and economic consequences.
A more detail project can be found in [my GitHub repo](https://github.com/jecordjotse/Weather_EventsAnalysis)

## REFERENCE {#refer}
[1] Reproducible Research Final Project in the John Hopkins University Data Science Specialization course on coursera <https://www.coursera.org/learn/reproducible-research/#syllabus>

## Apendix {#apendix}
#### 1- Content of the init.R file {#code}
```{r engine='bash', comment=''}
cat init.R
```

#### 2- EVTYPE {#evtype}
```{r}
levels(dataSub$EVTYPE)
```

#### 3- Heat Related Weather {#heatrelated}
```{r heatRel}
cat(levels(as.factor(heatRel)), sep = ", ")
```